### 2012-09-24 paper_sean/hyperneat.tex
made changes up to, but not including, the hyperneat section

% !TEX root =main.tex
HyperNEAT is a neuroevolution method that evolves artificial neuron networks (ANNs) using evolutionary algorithms.
HyperNEAT has shown to evolve large neural networks that represent many features present in brains such as repetitions and regularities \cite{stanley3}. 
This encoding has previously been applied for gait-learning \cite{yos:clune} and for evolving 2-D and 3-D objects \cite{clune:lipson} and has been successful in displaying complex, natural characteristics such as symmetries and repetition of themes often found in nature.

@@

>>>>

% !TEX root =main.tex
HyperNEAT is an algorithm for evolving artificial neural networks (ANNs)~\cite{stanley3}.

@@

Previous work has shown that Hyper
HyperNEAT has shown to evolve large neural networks that represent many features present in brains such as repetitions and regularities . 
This encoding has previously been applied for gait-learning \cite{yos:clune} and for evolving 2-D and 3-D objects \cite{clune:lipson} and has been successful in displaying complex, natural characteristics such as symmetries and repetition of themes often found in nature.

### 2012-09-24 paper_sean/introduction.tex
made changes up to, but not including, the hyperneat section

% !TEX root =main.tex
Robots that can move around on legs can operate in a much wider range of environments than their wheeled counterparts. A downside is that designing gaits for legged robots is a difficult and time-consuming process~\cite{ridder storm, wetter green}. 
But designing a gait for a robot is often difficult and time consuming because the engineer must take into consideration many factors which must all fall into place for a successful gait pattern. 
Because of setbacks present in hand-designing gaits, evolving gaits with machine learning algorithms is an attractive and flexible approach that can automatically produce effective gaits for many different robot platforms with minimal human intervention.

Much research has been done in this subject, and previous work has shown that learned gaits can outperform designed gaits \cite{valsalam:mii,kohl:stone,hornby1,hornby2}. %JMC/SL- I have a similar claim in my CEC 2009 paper. Grab those cites for here (especially the Hornby ones). / have added hornby citations
It has also been shown that gaits with regularity (coordination in leg movement, such as left-right symmetry or front-back symmetry) perform better than gaits without regularities \cite{valsalam:mii,clune2,clune1,clune3}. %JMC/SL-note how you can put multiple cites in one bracket...but no spaces after the comma! / gotcha!
However, in most previous works, the experimenter has had to manually decide the regularities of the gaits \cite{valsalam:mii,tellez,beer,raibert} . %JMC/SL - we need cites for this claim. you can get them from Clune1 / done
Such manual intervention is time consuming, requires expert knowledge, and adds constraints that may hurt performance. 
Previous work has shown that the Hypercube-based NeuroEvolution of Augmenting Topologies (HyperNEAT) generative encoding \cite{stanley1} can automatically generate a variety of regular gaits and that it outperforms direct encoding controls on this task \cite{clune1}, \cite{clune2}. 
However, that work only verified these claims in simulation. 
A follow-up paper by Yosinski et al. evolved gaits with HyperNEAT directly in hardware on the QuadraTot robot platform and found that HyperNEAT's gaits outperformed manually designed parameterized learning algorithms, but still did not produce impressive, natural gaits \cite{yos:clune}. 
A further follow-up study built a simulator for the QuadraTot robot platform to test whether the inclusion of a simulator would improve results and found that it did:  an evolutionary algorithm with direct encoding in simulation produced gaits that outperformed those gaits evolved by HyperNEAT in hardware \cite{glette}. %JMC/SL-Should there be a space between the end of the sentence and the cite? Whatever the answer, we should be consistent. If so, I have added here with the ~. The ~ tells Latex not to split that space from the character before and after it. / There should have been and will make changes accordingly
The simulator reduced the time required to evaluate gaits and afforded much larger population sizes and more generations than was possible when evolving in hardware, ultimately performing 333 times as many evaluations as HyperNEAT had in hardware per run (60000 vs. 180) \cite{yos:clune}, \cite{glette}. 
The evolutionary algorithm used in Glette et al. \cite{glette} was manually constrained to evolve specific, hand-chosen regularities. 
%Another study evolved gaits in hardware on the QuadraTot platform using RL PoWER, a spline-based algorithm, and was successful in evolving faster and more repeatable gaits than HyperNEAT evolved in hardware \cite{haocheng}. %JMC- I don't think we need to cite this paper...it is a non-evolutionary method, and it doesn't outperform Glette...and the sentence just sticks out as a detail that isn't necessary to our story.

In this paper we test whether the benefits of HyperNEAT -- higher performance and the automatic discovery of complex, effective regularities \cite{clune2,yos:clune} -- will outperform a simple genetic algorithm manually designed to produce regularities once HyperNEAT is combined with a simulator. %JMC/SL-I don't think they used a direct encoding. I prefer describing it as simple and manually designed to produce specific regularities. / changed to reflect that
We test this by evolving gaits using HyperNEAT with a physical simulator and then comparing the performance of these gaits in the real-world to those that were evaluated on the QuadraTot robotic platform in previous publications.  %JMC - Try to write without ever saying ``use'' or ``using.'' If you do, you'll find how much it is overused in writing, as evidenced by this last sentence. Often the word is a stand in for a different, clearer verb...and its use is ambiguous. Please search for this word and try to swap it out throughout the rest of the paper. Often times it requires restructuring a sentence, but the sentence usually ends up better. Similarly, make sure to always follow the word ``this'' with referent. This is extremely ambiguous. You know what you refer to when you write the sentence, but readers (often including your future self!) do not know precisely what you mean, and in science precision is critical. 
%Our experiments confirm the hypothesis: gaits evolved with HyperNEAT and then transferred to reality were the highest performing observed to date on the QuadraTot platform. 
%That said, gaits still transferred poorly, revealing that substantial further performance gains are likely via future work that combines a simulator, HyperNEAT, and techniques that reduce the simulator-to-reality gap. 
%More broadly, this work supports previous work that evolving with a simulator and then transferring to reality is a more powerful means of achieving results than evolving directly in hardware. 
%Our work also confirms that generative encodings such as HyperNEAT are a powerful way to evolve gaits and, more generally, automatically exploit regularities in challenging engineering domains.

>>>>

% !TEX root =main.tex
Robots that can move around on legs can operate in a much wider range of environments than their wheeled counterparts. A downside is that designing gaits for legged robots is a difficult and time-consuming process for human engineers~\cite{ridder storm, wetter green}, and must be repeated every time the design of a robot is changed or a new robot is created~\cite{hornby1}. 
Because of these difficulties in hand-designing gaits, scientists have investigated automatically producing gaits with machine learning or evolutionary algorithms, an approach that often produces better gaits than those created by human engineers~\cite{valsalam:mii,kohl:stone,hornby1,hornby2,yos:clune}. Both machine-created and human-created gaits perform better if they are \emph{regular}, in that they have coordinated movements, such as left-right symmetry or front-back symmetry~\cite{valsalam:mii,clune2,clune1,clune3}.
However, in most previous works, the experimenter has had to manually decide the regularities of the evolved gaits \cite{valsalam:mii,tellez,beer,raibert} . 
Such manual intervention is time consuming, requires expert knowledge, and adds constraints that may hurt performance.

Previous work has shown that the Hypercube-based NeuroEvolution of Augmenting Topologies (HyperNEAT) generative encoding \cite{stanley1} can automatically generate a variety of regular gaits and that it outperforms direct encoding controls on this task \cite{clune1}, \cite{clune2}. 
However, that work only verified these claims in simulation. 
A follow-up paper by Yosinski et al. evolved gaits with HyperNEAT directly in hardware on the QuadraTot robot platform and found that HyperNEAT's gaits outperformed manually designed parameterized learning algorithms, but still did not produce impressive, natural gaits \cite{yos:clune}. 
A further follow-up study built a simulator for the QuadraTot robot platform to test whether the inclusion of a simulator would improve results and found that it did:  when gaits were evolved in this simulator with a simple direct encoding and then transferred to the QuadraTot robot, the resulting gaits were faster than the gaits produced by evolving gaits with HyperNEAT directly on the robot \cite{glette}. 
The simulator helped because it afforded much larger population sizes and more generations than were possible when evolving directly in hardware. Specifically, the simulator allowed 333 times as many evaluations per run as HyperNEAT had in hardware (60000 vs. 180)~\cite{yos:clune}, \cite{glette}.

The work with the simulator (Glette et al.~\cite{glette}) evolved gaits with a simple encoding that was manually constrained to produce specific regularities. The success of that work raises the question of whether the performance gains were due to the use of a simulator, or the use of a simple, manually-constrained encoding. We hypothesized that HyperNEAT, which has been previously shown to automatically discover complex regularities to produce high-performing gaits~\cite{clune2,yos:clune}, would outperform the simpler encoding from Glette et al.~\cite{glette} if HyperNEAT is combined with a simulator. 
Here we test that hypothesis by evolving gaits with HyperNEAT in the same simulator from Glette et al. and then transferring those gaits to the QuadraTot robot.  Our experiments confirm the hypothesis: gaits evolved with HyperNEAT and then transferred to reality were the highest performing observed to date on the QuadraTot robot platform.

%The results supports previous work that evolving with a simulator and then transferring to reality is a more powerful means of achieving results than evolving directly in hardware. %JMC: Can you add a few cites here, hopefully from our current cite list? I think some of the JBM papers show this. 
%Our work also confirms that generative encodings such as HyperNEAT are a powerful way to evolve gaits and, more generally, automatically exploit regularities in challenging engineering domains.

### 2012-09-24 paper_sean/robot.tex
made changes up to, but not including, the hyperneat section

% !TEX root =main.tex
The robot platform used for this project was a quadrupedal robot named the QuadraTot\footnote{Robot plans and part files for 3D printing are freely available at http://quadratot.yosinski.com} designed in the Cornell Creative Machines Lab by Juan Zagal \cite{yos:clune}.
It has 9 degrees-of-freedom (DOF), with 2 DOF for the inner joint and the outer joint of each leg and 1 DOF for the hip joint of the robot. 
The QuadraTot is an open-sourced and 3-D printable robot, which allows it to be used in other labs for research, providing a platform with which people could compare the results of different gait learning algorithms. 
Currently, there are published data and results from 9 different learning algorithms on the platform from three different publications \cite{yos:clune,glette,haocheng}. %JMC/SL fill in INSERT, and dump all cites here. This is a good place to cite haochengs paper, which is fun for him and Jason, but does not distract from our story./ DONE

The body of the robot was printed using an Objet Connex 500 3-D Printing System and is actuated using 9 Robotis Dynamixel servos: 5 AX-18A servos and 4 AX-12A servos. 
The stronger AX-18A servos were used to actuate the inner joints of the robot and the hip joint while the AX-12A servos were used to actuate the outer joints. 
The servos have a range of motion of [0, 1023], roughly equivalent to a physical range of [-12\degree, +120\degree].

The Dynamixel AX-12A and AX-18A servos have a safety feature called \emph{Alarm Shutdown} that would turn off the servo if the servo's current, range, temperature, or torque was too high \cite{robotis}. 
The ones most relevant to this study were shutdowns due to high torque and range. 
It has been reported previously Yosinski et al. \cite{yos:clune}, that many gait tested on the QuadraTot forced the servos to exert too much torque or command positions that are out of range and cause the servos to shutdown. In order to address this problem and avoid collisions with its own body, we implemented a motion cropping system by limiting the possible range of movement of the QuadraTot by [-85\degree, +60\degree], [-113\degree, +39\degree], and [-28\degree, +28\degree] for the inner, outer, and hip joints respectively.
In addition to this, we implemented an extra cropping system named the Smart Cropping System from Shen et al. \cite{haocheng} to further prevent servo shutdowns from leg positions where the combination of the two angles interact to produce an extreme amount of torque. %This smart cropping algorithm calculates a leg position by summing the servo position values from the inner and outer joints of a leg, \emph{\textbf{g}} (\emph{\textbf{g}} = $j_{inner}$ + $j_{outer}$) %JMC/SL - do they sum the position values, or the joint angles? Since all the previous extremes are expressed in angles, the writing makes me expect angles... / They do use the position values and I think that it would be best to be consistent with what Jason/Haocheng did. This is also the way g is calculated in the simulator and the python drivers as well and more accurately reflects what we actually did. 
%and modifying the servo positions from ``fierce gaits'', which usually have relatively low \emph{\textbf{g}}s, to remap them onto the boundary of a quasi-trangular area in a 2-D space defined by $j_{inner}$ and $j_{outer}$. %JMC/SL - the term fierce sounds silly and non-professional (sorry!) I took out the place where you defined it, and we only use it in two other places. No need to invent a definition only to use it 2-3 times...can you swap in different, more descriptive words? / Again, I am trying to be consistent with what Jason and Haocheng used. I agree that 'fierce' is not the best word but if it is different from what J/H did, then there are inconsistencies between the publications...
%We set the minimum threshold \emph{\textbf{g}} to be 730, found through experimentation to provide a good balance between performance and safety. %JMC/SL - This description is not great. The last sentence's use of g doesn't make sense with the equation...is g a threshold or the sum? Please try to clean up this description. / Think this clarified things

We controlled the robot using a CompuLab Fit-PC2, a compact computer running Ubuntu Linux 10.10 and commanded the servos at 40Hz via the Pydynamixel library.

%In a previous study \cite{yos:clune}, this computer was placed on the robot to keep the robot untethered, but for this study the computer was taken off the robot to reduce weight, because the weight of the robot seemed to be one of the factors contributing to servo shutdowns.

We identified the robot's position using infrared LEDs on the robot which was tracked using a Wiimote as in Yosinski et al. \cite{yos:clune}.
%JMC/SL you sure there was more than one IR LED? I thought it was just one IR and one visible for us to tell if there is power to the LEDs. If it is just one, switch to singular instead of plural in the previous sentence / Yep there are like 3 IR LEDs being used.
%To maximize its viewable window, the Wii remote was fixed to the ceiling pointing down. It communicated with the robot computer via bluetooth using the CWiid library \cite{yos:clune}. 
%JMC/SL The paragraph about the Wii mote was very wordy. I think I trimmed it by half. You should try to be similarly compact the rest of the paper. Compare your version to mine and try to see how I condensed things. / Will do boss.

% % figures
% % figure of the quadratot in its natural stance
% \begin{figure}
% \begin{center}
% \vspace{1cm}
% % just using a photo from Haocheng's paper! Should replace if needed be           
% \epsfig{file=quadratot_fig.eps, width=8cm}
% \caption[ ]{The QuadraTot robotic platform. It is open-sourced and 3-D printable, which has allowed for other researchers to use the robot for their research around the world. The robot has 9 DOF, 2 per leg and a hip joint which connects the two symmetrical halves of the robot. The robot uses Dynamixel AX-18 and AX-12 servos for actuation. Figure taken from Shen et al. \cite{haocheng}.}
% \label{fig:quadratot}
% \end{center}
% \end{figure}
% %

%It is open-sourced and 3-D printable, which has allowed for other  researchers to use the robot for their research around the  world. 
%JMC/SL The previous sentence (which I commented out) is alright for the main text, but figures are very expensive real estate. The goal of our paper is not to sell this robot, but to sell the work we did. / Okay I understand

%JMC/SL: If you use which instead of that, you need a comma before which. For example, this is wrong: There are 9 degrees of freedom: 2 per leg, and a single `hip joint' which connects  the two symmetrical halves of the robot. This is right. There are 9 degrees of freedom: 2 per leg, and a single `hip joint' that connects  the two symmetrical halves of the robot. This is also right: There are 9 degrees of freedom: 2 per leg, and a single `hip joint,' which connects  the two symmetrical halves of the robot. / Ok
% The robot uses Dynamixel  AX-18 and AX-12 servos for actuation. %JMC We said this already in methods
% Figure taken from Shen et  al. \cite{haocheng}.%JMC Actually it's from yos:clune, but I think it is ok to leave this out, since this is just a stock photo of the QuadraTot.
 %JMC Let's put this figure on the first page (before the introduction), unless you can get a series of pictures that show a gait. (This might be hard to do). A new trend in papers is to have a sexy image on the first page, and Hod will want to see this. I also think it makes people want to read the paper more.

% \begin{figure}
% \begin{center}
% \vspace{1cm}
% \epsfig{file=legpos.eps, width=8cm}
% \caption[ ]{The figure shows an example of leg positions allowed and not allowed on the robot. The twoleft positions are allowed. The middle two positions are on the borderline. The right two positions are considered as ``fierce'' and are disallowed. Instead, these gaits are cropped to the middle two positions during gait performance by the cropping mechanism placed to prevent servo shutdowns. Figure taken from Shen et al. \cite{haocheng}}
% \end{center}
% \end{figure}
\figp{sixLegPositions}{.6}{The figure shows an example of leg positions allowed and not allowed on the robot. The twoleft positions are allowed. The middle two positions are on the borderline. The right two positions are considered as ``fierce'' and are disallowed. Instead, these gaits are cropped to the middle two positions during gait performance by the cropping mechanism placed to prevent servo shutdowns. Figure taken from Shen et al. \cite{haocheng}}

>>>>

% !TEX root =main.tex
We performed the experiments on the QuadraTot quadrupedal robot platform~\cite{yos:clune}.
It has 9 degrees-of-freedom (DOF): two joints per leg and one joint that allowed rotation along the robot's midline~(Fig.\ref{pictureOfRobot}). %we def want a picture of the cool robot. I propose putting the picture of the robot on the left and the picture of the simulated robot on the right of the same figure. That also does a nice job of highlighting the theme of the paper (robot+sim) in one figure. Try to get this figure to show up as early as possible in the paper (e.g top of the 2nd page). 
QuadraTot has open source software and hardware\footnote{A list of all of the parts and all software, including the simulator, is available for free at http://quadratot.yosinski.com.} %JMC: This should really be a Creative Machines/Cornell web address. Can you ask Jason if he agrees and, if so, take the lead on making sure it gets made?
, including many parts that can be 3D printed. These traits enable other labs to perform research with it, providing a platform for comparing different gait-learning algorithms. 
To date there are results on the platform from nine different learning algorithms from three previous publications \cite{yos:clune,glette,haocheng}. We identified the robot's position using infrared LEDs on the robot which was tracked using a Wiimote~\cite{yos:clune}.

Both in this work and that of Glette et al.~\cite{glette}, the joints are powered by Robotis Dynamixel servos; five AX-18A servos for the inner joint of each leg and the single midline joint, and four AX-12A servos for the the outer joint of each leg, which requires less power and can thus have less expensive motors. The servos were sent new positions at 40Hz via the Pydynamixel library.

Each servo has a built-in safety mechanism that shuts itself off to prevent damage if the servo's current, range, temperature, or torque was too high \cite{robotis}. During evolution, this safety mechanism frequently activated, and did so inconsistently, adding significant noise to the evaluation process. As pointed out in Yosinski et al. \cite{yos:clune}, gaits generated during evolution are highly variable and generate many shutdowns because they forced the servos to exert too much torque or command positions that are out of range. %JMC: Does requesting a value out of range really cause the servo to shut down? That seems like something we could easily fix (by never allowing an out-of-range value). Now that we do not allow values out of range, we still see lots of shutdowns...making me think that the problem was really torque. What do you think?
To minimize shutdowns, we limited the allowable range of movement for the inner, outer, and hip joints to [-85\degree, +60\degree], [-113\degree, +39\degree], and [-28\degree, +28\degree], respectively. We also implemented the Smart Cropping System from Shen et al. \cite{haocheng}, which prevents combinations of joint positions for the inner and outer joint of each leg that generate extreme amounts of torque. A final method of reducing torque was to reduce the weight of the robot. Yosinski et al.~\cite{yos:clune} had the small Linux computer that performed all computation on the robot, but we removed it and sent commands to the robot via a cable.

% % figure of the quadratot in its natural stance
% \begin{figure}
% \begin{center}
% \vspace{1cm}
% % just using a photo from Haocheng's paper! Should replace if needed be           
% \epsfig{file=quadratot_fig.eps, width=8cm}
% \caption[ ]{The QuadraTot robotic platform. It is open-sourced and 3-D printable, which has allowed for other researchers to use the robot for their research around the world. The robot has 9 DOF, 2 per leg and a hip joint which connects the two symmetrical halves of the robot. The robot uses Dynamixel AX-18 and AX-12 servos for actuation. Figure taken from Shen et al. \cite{haocheng}.}
% \label{fig:quadratot}
% \end{center}
% \end{figure}
% %

\figp{sixLegPositions}{.6}{Allowed and disallowed leg position angles. %JMC: Isn't this described in Haocheng's paper? I don't think we need to show this here, because it is not a new thing in our paper and is not central to our story. If we do keep it, instead of 3 sentences of text saying what the columns are, just put the headings ``Allowed'', ``End of Allowable Range'', and ``Disallowed''
%The figure shows an example of leg positions allowed and not allowed on the robot. The two left positions are allowed. The middle two positions are on the borderline. The right two positions are considered as ``fierce'' and are disallowed. Instead, these 
Disallowed gaits are cropped to the middle two positions to prevent servo shutdowns. Figure taken from Shen et al. \cite{haocheng}}

### 2012-09-24 paper_sean/simulator.tex
made changes up to, but not including, the hyperneat section

% !TEX root =main.tex
Simulations occurred in C++ software customized for the QuadraTot\footnote{Available for Linux and Windows at http://quadratot.yosinski.com} in the NVIDIA PhysX physics simulator, as described previously \cite{glette}. 
Specifically, the simulation captured the number and type of joints, mass, and rough shapes and lengths of the QuadraTot components. 
%The simulator was visualized by OpenGL library to allow the user to have a more intuitive grasp of what is going on. %JMC Unnecessary detail since this paper is not about the simulator. 
%The simulator visualization included a recording feature which allows the user to create videos of the robot in simulation.%JMC: Assumed
%The simulator also features a command line function which allows the users to input a gait file and run it on the simulator. %JMC: Assumed
%The simulator then outputs a file containing the X, Y, Z coordinates of the center of mass of the robot over time. %JMC: Assumed
%In simulation, the robot traverses in the XY plane with the Z axis representing the height off the ground. %JMC Unnecessary detail
In an attempt to more accurately model the robot behavior, we implemented the same cropping function in the simulator as in the actual robot. %JMC/SL - Do all results in this paper reflect this? / They will soon.
Although the simulator was originally written in Windows for the Windows operating platform we ported and ran it on Linux (Ubuntu 11.04) to seamlessly link the simulator and the already exising libraries together. 
The friction values used for this study was the same as that used in Glette et al. \cite{glette}. 
%
% and uses OpenGL and GLFW libraries for visualization. The simula- tor was originally written for the Windows operating system but was ported and run on Linux (Ubuntu 11.04) for this ex- periment. It features a commandline function which allows the user to specify a gait input file and extract an output file with the X, Y, Z coordinate information over time. The Eu- clidean distance travelled by the robot in the XY plane was used to calculate the speed of the robot in simulation and was used as the fitness value used for evolution.
%

% \begin{figure}
% \begin{center}
% \vspace{1cm}
% \epsfig{file=simulator.eps, width=8cm}
% \caption[ ]{Physical representation of the QuadraTot in simulation. It captures the important parts of the robot, such as the number and types of the joints, mass, and the rough shape and lengths. The simulator was written by Kyrre Glette in C++ using NVIDIA PhysX software library and visualized using OpenGL. Figure taken from Glette et al. \cite{glette}.}
% \end{center}
% \end{figure}
\figp{simulator_outline}{.6}{Physical representation of the QuadraTot in simulation. It captures the important parts of the robot, such as the number and types of the joints, mass, and the rough shape and lengths. The simulator was written by Kyrre Glette in C++ using NVIDIA PhysX software library and visualized using OpenGL. Figure taken from Glette et al. \cite{glette}.}

>>>>

% !TEX root =main.tex
The simulator\footnote{Available for Linux and Windows at http://quadratot.yosinski.com} was the same one used in Glette et al. \cite{glette}. C++ code for the the Nvidia PhysX physics engine approximately describes the QuadraTot, including the mass and size of the robot components and its degrees of freedom. In the simulator, each individual joint range was limit as described above, but Smart Cropping was not included because it hindered performance by limiting the types of gaits evolution could explore in early generations.

\figp{simulator_outline}{.6}{Physical representation of the QuadraTot in simulation. It captures the important parts of the robot, such as the number and types of the joints, mass, and the rough shape and lengths. The simulator was written by Kyrre Glette in C++ using NVIDIA PhysX software library and visualized using OpenGL. Figure taken from Glette et al. \cite{glette}.}
